TODO LIST

### 2019-01-22 LMCS/sendrev/response-letter.tex

	A proof was added.

%	\item[$\checkmark$] p.31, definition of $\leq$ on $\F$: ``$f(i) \leq g(i)$ for every $i \in \{1,\ldots,\ell\}$'': Inputs for $f$ and $g$ should be elements of $A^{\ell}$, not $\{1,\ldots,\ell\}$
	\item \red{p.32/33, example 6.2: I don't see why the formula given for dag's does not work for arbitrary graphs: In arbitrary graphs, the fixpoint is still reached after at most $n-1$ steps, since the maximal length of any shortest path in any graph is $n-1$. On the other hand, assuming that I am mistaken and the support does not become stable within $n-1$ steps: Using the formula	given for arbitrary graphs, the support of the constructed function restricted to those tuples that have first in their last component is the same as the support of the constructed function using the formula for dags in each step of the recursion. Hence, the formula should not be able to fix such a problem.}

	The problem with the first formula for directed graphs is precisely the opposite. The fixed point could be reached {\em before} we want it to. The fixed point is always reached after $k$ steps where $k$ is the size of the maximum shortest path between two vertices in the graph, and if the graph has a cycle in it, there might be paths of size $k' \in [k+1,n]$ that wouldn't be counted because the fixed point was reached earlier.

%	In a graph that is only one cycle of size $n$, namely $C_n$, the fixed point is reached after step $n$.

@@

\begin{itemize}
	\setlength\itemsep{0.5em}
	\item In several proofs, you rely on disjoint union of solutions and cartesian product of solutions. Indeed, its gives a polynomial time counting reduction. Can you make connections with (+,*) circuits ? It would be similar to what is done in A Circuit-Based Approach to Efficient Enumeration (Antoine Amarilli, Pierre Bourhis, Louis Jachiet, Stefan Mengel) where simple circuits (dDNNF, which are basically disjoint unions and cartesian products) are used to capture the solutions of MSO queries.

	\item In general I would be interested to know the relationship of your couting classes with their enumeration counterpart. 
	For instance there is a paper adapting [39] to enumeration: Enumeration Complexity of Logical Query Problems with Second-order Variables (Arnaud Durand and Yann Strozecki). In particular they consider class with some Horn structure to capture easy enumeration problems, it may have some connection with the class $\sigma$ QSO($\Pi_1$ Horn) you introduce.

	\item For the most involved proofs, you could describe informally your proof strategy in a few lines to help the reader. You may also add some example to illustrate the different fragment of QSO (you already do it sometimes and it is very helpful).

%	\item[$\checkmark$] page 4: -some of the $\to$ some of them 
%	\item[$\checkmark$] page 10:  -$2^{|A|^l}$-bits $\to$ $2^{|A|^l}$ bits
%	\item[$\checkmark$] page 21: -it does not mentioned $\to$ it does not mention
%	\item page 22: -you make a long proof about some closure under subtraction by one. You should put the result in a theorem envirpnment to make what you are proving more evident.
%	\item[$\checkmark$] -replace .... for $\to$ replace ... by;  there are several instances of that
%	\item[$\checkmark$] -$\phi$ is an FO-formula $\to$ a $\Sigma_1[FO]$ formula
%	\item -you explain that you rewrite your formula to obtain conditions a,b and c, you should give an idea how these conditions will be used
%	\item[$\checkmark$] page 23: a SO, a FO  an SO, an FO
%	\item page 24: 	if, if $\phi$$\to$ $\phi$
%	\item page 25: 	-you explain that the proof relies on witnesses of logarithmic size.	It was not clear to me while reading the proof. Can you explain more ?
	\item page 28:	-the complete problem for $\Sigma$ QSO($\Sigma_2$ Horn) would be complete also for $\#\Sigma-2$ Horn. You should mention here that you do not need sum here	and that the framework of [39] is enough for this result.
%	\item[?] page 30: ``this notion can be used to captures FP'' $\to$ ``... to capture FP''
%	\item[$\checkmark$] page 32: - paths of length at most n $\to$ in an acylic graph, all paths are of size less than n
%	\item page 33: 	-there is a proof symbol \verb|\qed| but it does not end a well defined proof environment
	\item page 38: the last sentence on fixed point operator is to vague. Either expand it to say what king of question on fixed point operator you want to investigate or remove it	
\end{itemize}

### 2019-01-22 LMCS/sendrev/proof-sigmafo-minusone.tex
correcciones Reviewer 2

The proof will be separated in two parts. In the first part, we will assume that  $\alpha$ is in $\logex{1}$-PNF, namely, $\alpha = \sa{\bar{X}}\sa{\bar{x}}\varphi$ for some $\varphi$ in $\logex{1}$. We will show how to define a formula $\varphi'$ that satisfies the following condition: for each $\A$, if $(\A,V,v)\models \varphi(\bar{X},\bar{x})$ for some $V$ and $v$ over $\A$, then there exists exactly one assignment to $(\bar{X},\bar{x})$ that satisfies $\varphi$ and not $\varphi'$. 
From this, we will have that $\kappa(\alpha) = \sa{\bar{X}}\sa{\bar{x}}\varphi'$ is the desired formula. In the second part, we suppose that $\alpha$ is of the form $\beta + \sa{\bar{X}}\sa{\bar{x}}\varphi$ with $\beta$ being the sum of one or more formulae in $\logex{1}$-PNF. 
We define a formula $\varphi'$ that satisfies the following condition: if $(\A,V,v)\models \varphi(\bar{X},\bar{x})$ and $\sem{\beta}(\A) = 0$, then there exists exactly one assignment to $(\bar{X},\bar{x})$ that satisfies $\varphi$ and not $\varphi'$. From here, we can define $\kappa$ recursively as  $\kappa(\alpha) = \kappa(\beta) +  \sa{\bar{X}}\sa{\bar{x}}\varphi'$ and the property of subtraction by one will be~proven.

@@

To simplify the analysis of $\varphi$, the first step is to rewrite $\varphi$ in a DNF formula. 
More precisely, we rewrite $\varphi$ into an equivalent formula of the form $\bigvee_{i = 1}^m \varphi_i$ for some $m\in \bbN$ where each $\varphi_i(\bar{X}, \bar{x}) = \ex{\bar{y}} \varphi_i'(\bar{X}, \bar{x}, \bar{y})$ and $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ is a conjunction of atomic formulae or negation of atomic formulae. Furthermore, we assume that each $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ has the~form:
$$
\varphi_i'(\bar{X}, \bar{x}, \bar{y}) =  \underbrace{\varphi_i^{\fo}(\bar{x},\bar{y})}_{\text{an $\fo$ formula}} \wedge 
\underbrace{\varphi_i^{+}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $X_j$'s}} \wedge
\underbrace{\varphi_i^{-}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $\neg X_j$'s}}.
$$
Note that atomic formulae, like $R(\bar{z})$ for $R \in \R$, will appear in the subformula $\varphi_i^{\fo}(\bar{x},\bar{y})$.

Now, we define a series of rewritings to $\varphi$ that will make each formula $\varphi_i$ satisfy the following three conditions: (a) no variable from $\bar{x}$ appears in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$, (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines a weak ordering over the variables in $(\bar{x},\bar{y})$ (the precise definition of a weak ordering is detailed below) and (c) if both $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ appear in the formula, the weak ordering should not satisfy $\bar{z} = \bar{w}$.
We explain below how to rewrite $\varphi_i$ in order to satisfy each condition.

@@

\noindent {\em (a) No variable from $\bar{x}$ appears in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$.} In order to satisfy this condition, consider some instance of $X_j(\bar{w})$ in $\varphi_i$, where $\bar{w}$ is a subtuple of $(\bar{x},\bar{y})$. Add $\length{\bar{w}}$ new variables $z_1,\ldots,z_{\length{\bar{w}}}$ to the formula and let $\bar{z} = (z_1,\ldots,z_{\length{\bar{w}}})$. We redefine $\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$ by replacing $X_j(\bar{w})$ with $X_j(\bar{z})$ (denoted by $\varphi_i^{+}(\bar{X},\bar{x},\bar{y})[X_j(\bar{w}) \leftarrow X_j(\bar{z})]$ and then the formula $\varphi_i$ is equivalently defined as:
$$
\varphi_i(\bar{X},\bar{x}) \ := \ \ex{\bar{y}} \ex{\bar{z}} \big( \bar{z} = \bar{w} \wedge \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{x},\bar{y})[X_j(\bar{w}) \leftarrow X_j(\bar{z})] \wedge
\varphi_i^{-}(\bar{X},\bar{x},\bar{y}) \big).
$$
We repeat this process for each instance of a $X_j(\bar{w})$ in $\varphi_i$, and we obtain a formula where none of the $X_j$'s acts over any variable in $\bar{x}$. We add the new first-order variables to $\bar{y}$ and we redefine $\varphi_i$ as:
$$
\varphi_i(\bar{X},\bar{x}) \ := \  \ex{\bar{y}} \big( \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y})\big).
$$
For example, if $\bar{x} = x$, $\bar{y} = y$ and $\varphi_i = \ex{\bar{y}} \big( x < y \wedge  X(x,y)\wedge \neg X(x,x)\big)$, then we redefine $\bar{y} = (y,v_1,v_2,v_3,v_4)$ and:
$$
\varphi_i \ := \ \ex{\bar{y}}\big( v_1 = x \wedge v_2 = y \wedge v_3 = x \wedge v_4 = x \wedge x < y  \wedge  X(v_1,v_2) \wedge \neg X(v_3,v_4)\big).
$$
\noindent {\em (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines a weak ordering over the variables in $(\bar{x},\bar{y})$.} A weak ordering on a set $S$ is defined by an equivalence relation $\sim$ over $S$, and a linear order over $S/\!\sim$. For example, let $\bar{x} = (x_1,x_2,x_3,x_4)$. A possible weak ordering would be defined by the formula $\theta(\bar{x}) = x_2 < x_1 \wedge x_1 = x_4 \wedge x_4 < x_3$. On the other hand, the formula $\theta'(\bar{x}) = x_1 < x_2 \wedge x_1 < x_4 \wedge x_2 = x_3$ does not define a weak ordering since both $\{x_1\}<\{x_2,x_3\}<\{x_4\}$ and $\{x_1\} < \{x_2,x_3,x_4\}$ satisfy $\theta'$.
For a given $k$, let $\cB_k$ be the number of possible weak orderings for a set of size $k$. For $1 \leq j \leq \cB_{\length{(\bar{x},\bar{y})}}$ 
let $\theta^j(\bar{x},\bar{y})$ be the formula that defines the $j$-th weak ordering over $(\bar{x},\bar{y})$. 
The formula $\varphi(\bar{X},\bar{x})$ is thus redefined as:
$$
\varphi(\bar{X},\bar{x}) \ := \ \bigvee_{i = 1}^m \bigvee_{j = 1}^{\cB_{\length{(\bar{x},\bar{y})}}} \ex{\bar{y}} \big(\theta^j(\bar{x},\bar{y})\wedge \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y})\big),
$$
Note that each $\theta^j(\bar{x},\bar{y})$ is an $\fo$-formula.
Then, by redefining $\varphi_i^{\fo}(\bar{x},\bar{y})$  as $\theta^j(\bar{x},\bar{y})\wedge \varphi_i^{\fo}(\bar{x},\bar{y})$, we can suppose that each $\varphi_i^{\fo}(\bar{x},\bar{y})$ forces a weak ordering over the variables in $(\bar{x},\bar{y})$.

\noindent {\em (c) If both $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ appear in the formula, the weak ordering should not satisfy $\bar{z} = \bar{w}$.} 
%Finally, to rewrite the formula such that no $X_j(\bar{z})$ or $\neg X_j(\bar{w})$ appear in $\varphi_i$ while $\bar{z}$ and $\bar{w}$ are made equal in the weak ordering (which would make $\varphi_i$ inconsistent), we do the following.
If there exists an instance of $X_j(\bar{z})$ in $\varphi^{+}_i$, an instance of $\neg X_j(\bar{w})$ in $\varphi^{-}_i$ and the weak ordering in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$, then the entire formula $\varphi_i$ is removed from $\varphi$.
It is important to notice that the resulting $\varphi$ is equivalent to the initial one, and it is still a formula in $\eqso(\logex{1})$. From now on, we assume that each $\varphi_i(\bar{X},\bar{x}) = \ex{\bar{y}}  \varphi_i'(\bar{X},\bar{x}, \bar{y}) $ satisfies conditions (a), (b) and (c), and $\varphi_i'(\bar{X},\bar{x}, \bar{y})$ has the~form:
$$
\varphi_i'(\bar{X},\bar{x}, \bar{y}) \; = \; \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y}) 
$$
where $\varphi^{+}$ and $\varphi^{-}_i$ do not depend on $\bar{x}$.
\begin{clm}\label{claim:minusone}
	For an ordered structure $\A$ and an FO assignment $v$ for $\A$, $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$ if, and only if, there exists an SO assignment $V$ for $\A$ such that $(\A,V,v)\models \varphi_i'(\bar{X},\bar{x}, \bar{y})$.
\end{clm}
\proof
	To prove the {\em only if} direction, let $\A$ be an ordered structure with domain $A$ and let $v$ be a first-order assignment for $\A$, such that $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$.
	Define $\bar{B} = (B_1,\ldots,B_{\length{\bar{X}}})$ as $B_j = \{v(\bar{w})\mid \text{ $X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$, and let $V$ be a second-order assignment for which $V(\bar{X}) = \bar{B}$.
	Towards a contradiction, suppose that $(\A,V,v)\not\models \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$.
	By the choice of $v$, and construction of $V$ it is clear that $(\A,V,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})\wedge\varphi^{+}_i(\bar{X},\bar{y})$, so we necessarily have that $(\A,V,v)\not\models\varphi^{-}_i(\bar{X},\bar{y})$.
	Let $X_t$ be such that $\neg X_t(\bar{w})$ is mentioned in $\varphi^{-}_i(\bar{X},\bar{y})$ and $(\A,V, v)\not\models\neg X_t(\bar{w})$, namely, $v(\bar{w})\in B_t$. 
	However, by the construction of $B_t$, there exists a subtuple $\bar{z}$ of $\bar{y}$ such that $X_t(\bar{z})$ appears in $\varphi^{+}_i(\bar{X},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$. Since $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$, then the weak ordering in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$. This violates condition (c) above since $\neg X_t(\bar{w})$ appears in $\varphi^{-}_i$ and $X_t(\bar{z})$ appears in $\varphi^{+}_i$, which leads to a contradiction.

@@

The previous claim and proof motivate the following definition.
For a structure $\A$ and a first-order assignment $v$ for $\A$, define $\bar{B}^v = (B^v_1,\ldots,B^v_{\length{\bar{X}}})$ where each $B^v_j = \{v(\bar{w}) \mid \text{$X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$.
One can easily check that for every assignment $(V, v)$ such that $(\A,V,v)\models \varphi_i'(\bar{X}, \bar{x}, \bar{y})$, it holds that $(\A,\bar{B}^v,v)\models \varphi_i'(\bar{X}, \bar{x}, \bar{y})$ and $\bar{B}^v \subseteq V(\bar{X})$, namely, $\bar{B}^v$ is a valid candidate for $\bar{X}$ and, furthermore, it is contained in all assignments of $\bar{X}$ when $v$ is fixed.
This offers an insight into the main idea of Part (1): by choosing one particular $v$ the plan is to remove $\bar{B}^v$ as an assignment over $\bar{X}$ in $\varphi_i$. 
For this, we choose the minimal $v$ that satisfies $\varphi^{\fo}_i(\bar{x},\bar{y})$ which can be defined with the following formula:
\[
\text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \; = \;  \,\varphi_i^{\fo}(\bar{x},\bar{y})\wedge \fa{\bar{x}'} \fa{\bar{y}'}\big(\varphi_i^{\fo}(\bar{x}',\bar{y}')\to ( \bar{x}\leq\bar{x}' \wedge  \bar{y}\leq\bar{y}')\big).
\] 
If $\varphi^{\fo}_i$ is satisfiable, let $v$ be the only assignment such that  $(\A,v)\models \text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y})$. 
Furthermore, let $V^*$ be the second-order assignment and $v^*$ the first-order assignment that satisfy $V^*(\bar{X}) = \bar{B}^{v}$ and $v^*(\bar{x}) = v(\bar{x})$.
By the previous discussion, $(\A,V^*,v^*)\models\varphi_i(\bar{X},\bar{x})$.

>>>>

{\em Proof outline:} The proof will be separated in two parts. In the first part, we will assume that  $\alpha$ is in $\logex{1}$-PNF, namely, $\alpha = \sa{\bar{X}}\sa{\bar{x}}\varphi$ for some $\varphi$ in $\logex{1}$. We will show how to define a formula $\varphi'$ that satisfies the following condition: for each $\A$, if $(\A,V,v)\models \varphi(\bar{X},\bar{x})$ for some $V$ and $v$ over $\A$, then there exists exactly one assignment to $(\bar{X},\bar{x})$ that satisfies $\varphi$ and not $\varphi'$. 
From this, we will have that $\kappa(\alpha) = \sa{\bar{X}}\sa{\bar{x}}\varphi'$ is the desired formula. In the second part, we suppose that $\alpha$ is of the form $\beta + \sa{\bar{X}}\sa{\bar{x}}\varphi$ with $\beta$ being the sum of one or more formulae in $\logex{1}$-PNF. 
We define a formula $\varphi'$ that satisfies the following condition: if $(\A,V,v)\models \varphi(\bar{X},\bar{x})$ and $\sem{\beta}(\A) = 0$, then there exists exactly one assignment to $(\bar{X},\bar{x})$ that satisfies $\varphi$ and not $\varphi'$. From here, we can define $\kappa$ recursively as  $\kappa(\alpha) = \kappa(\beta) +  \sa{\bar{X}}\sa{\bar{x}}\varphi'$ and the property of subtraction by one will be~proven.

@@

To simplify the analysis of $\varphi$, the first step is to rewrite $\varphi$ as a DNF formula. 
More precisely, we rewrite $\varphi$ into an equivalent formula of the form $\bigvee_{i = 1}^m \varphi_i$ for some $m\in \bbN$ where each $\varphi_i(\bar{X}, \bar{x}) = \ex{\bar{y}} \varphi_i'(\bar{X}, \bar{x}, \bar{y})$ and $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ is a conjunction of atomic formulae or negation of atomic formulae. Furthermore, we assume that each $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ has the~form:
$$
\varphi_i'(\bar{X}, \bar{x}, \bar{y}) =  \underbrace{\varphi_i^{\fo}(\bar{x},\bar{y})}_{\text{an $\fo$ formula}} \wedge 
\underbrace{\varphi_i^{+}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $X_j$'s}} \wedge
\underbrace{\varphi_i^{-}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $\neg X_j$'s}}.
$$
Note that atomic formulae, like $R(\bar{z})$ for $R \in \R$, will appear in the subformula $\varphi_i^{\fo}(\bar{x},\bar{y})$.

Now, we define a series of rewrites of $\varphi$ that will make each formula $\varphi_i$ satisfy the following three conditions: (a) no variable from $\bar{x}$ appears in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$, (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines a weak ordering over the variables in $(\bar{x},\bar{y})$ (the precise definition of a weak ordering is detailed below) and (c) if both $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ appear in the formula, the weak ordering should not satisfy $\bar{z} = \bar{w}$.
When these conditions are met, $\varphi_i'(\bar{X}, \bar{x}, \bar{y})$ will be satisfiable if and only if $\varphi_i^{\fo}(\bar{x},\bar{y})$ is satisfiable. This is proven in Claim~\ref{claim:minusone}.
We explain below how to rewrite $\varphi_i$ in order to satisfy each condition.
\martin{Peque\~na explicaci\'on de para qu\'e sirven las condiciones a b y c}

@@

\noindent {\em (a) No variable from $\bar{x}$ appears in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$.} In order to satisfy this condition, consider some instance of $X_j(\bar{w})$ in $\varphi_i$, where $\bar{w}$ is a subtuple of $(\bar{x},\bar{y})$. Add $\length{\bar{w}}$ new variables $z_1,\ldots,z_{\length{\bar{w}}}$ to the formula and let $\bar{z} = (z_1,\ldots,z_{\length{\bar{w}}})$. We rewrite $\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$ by replacing $X_j(\bar{w})$ with $X_j(\bar{z})$ (denoted by $\varphi_i^{+}(\bar{X},\bar{x},\bar{y})[X_j(\bar{w}) \leftarrow X_j(\bar{z})]$) and then the formula $\varphi_i$ is equivalently defined as:
$$
\varphi_i(\bar{X},\bar{x}) \ := \ \ex{\bar{y}} \ex{\bar{z}} \big( \bar{z} = \bar{w} \wedge \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{x},\bar{y})[X_j(\bar{w}) \leftarrow X_j(\bar{z})] \wedge
\varphi_i^{-}(\bar{X},\bar{x},\bar{y}) \big).
$$
We repeat this process for each instance of a $X_j(\bar{w})$ in $\varphi_i$, and we obtain a formula where none of the $X_j$'s acts over any variable in $\bar{x}$. We add the new first-order variables to $\bar{y}$ and we redefine $\varphi_i$ as:
$$
\varphi_i(\bar{X},\bar{x}) \ := \  \ex{\bar{y}} \big( \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y})\big).
$$
For example, if $\bar{x} = x$, $\bar{y} = y$ and $\varphi_i = \ex{\bar{y}} \big( x < y \wedge  X(x,y)\wedge \neg X(x,x)\big)$, then we redefine $\bar{y} = (y,v_1,v_2,v_3,v_4)$ and:
$$
\varphi_i \ := \ \ex{\bar{y}}\big( v_1 = x \wedge v_2 = y \wedge v_3 = x \wedge v_4 = x \wedge x < y  \wedge  X(v_1,v_2) \wedge \neg X(v_3,v_4)\big).
$$
\noindent {\em (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines a weak ordering over the variables in $(\bar{x},\bar{y})$.} A weak ordering on a set $S$ is defined by an equivalence relation $\sim$ over $S$, and a linear order over $S/\!\sim$. For example, let $\bar{x} = (x_1,x_2,x_3,x_4)$. A possible weak ordering would be defined by the formula $\theta(\bar{x}) = x_2 < x_1 \wedge x_1 = x_4 \wedge x_4 < x_3$. On the other hand, the formula $\theta'(\bar{x}) = x_1 < x_2 \wedge x_1 < x_4 \wedge x_2 = x_3$ does not define a weak ordering since both $\{x_1\}<\{x_2,x_3\}<\{x_4\}$ and $\{x_1\} < \{x_2,x_3,x_4\}$ satisfy $\theta'$.
For a given $k$, let $\cB_k$ be the number of possible weak orderings for a set of size $k$. For $1 \leq j \leq \cB_{\length{(\bar{x},\bar{y})}}$ 
let $\theta^j(\bar{x},\bar{y})$ be the formula that defines the $j$-th weak ordering over $(\bar{x},\bar{y})$. 
The formula $\varphi(\bar{X},\bar{x})$ is thus redefined as:
$$
\varphi(\bar{X},\bar{x}) \ := \ \bigvee_{i = 1}^m \bigvee_{j = 1}^{\cB_{\length{(\bar{x},\bar{y})}}} \ex{\bar{y}} \big(\theta^j(\bar{x},\bar{y})\wedge \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y})\big),
$$
Note that each $\theta^j(\bar{x},\bar{y})$ is an $\fo$-formula.
Then, by redefining $\varphi_i^{\fo}(\bar{x},\bar{y})$  as $\theta^j(\bar{x},\bar{y})\wedge \varphi_i^{\fo}(\bar{x},\bar{y})$, we can suppose that each $\varphi_i^{\fo}(\bar{x},\bar{y})$ forces a weak ordering over the variables in $(\bar{x},\bar{y})$.

\noindent {\em (c) If both $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ appear in the formula, the weak ordering should not satisfy $\bar{z} = \bar{w}$.} 
%Finally, to rewrite the formula such that no $X_j(\bar{z})$ or $\neg X_j(\bar{w})$ appear in $\varphi_i$ while $\bar{z}$ and $\bar{w}$ are made equal in the weak ordering (which would make $\varphi_i$ inconsistent), we do the following.
If there exists an instance of $X_j(\bar{z})$ in $\varphi^{+}_i$, an instance of $\neg X_j(\bar{w})$ in $\varphi^{-}_i$ and the weak ordering in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$, then the entire formula $\varphi_i$ is removed from $\varphi$.
It is important to notice that the resulting $\varphi$ is equivalent to the initial one, and it is still a formula in $\eqso(\logex{1})$. From now on, we assume that each $\varphi_i(\bar{X},\bar{x}) = \ex{\bar{y}}  \varphi_i'(\bar{X},\bar{x}, \bar{y}) $ satisfies conditions (a), (b) and (c), and moreover, $\varphi_i'(\bar{X},\bar{x}, \bar{y})$ has the~form:
$$
\varphi_i'(\bar{X},\bar{x}, \bar{y}) \; = \; \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y}).
$$
Note that neither one of $\varphi^{+}$ or $\varphi^{-}_i$ depends on $\bar{x}$.
\begin{clm}\label{claim:minusone}
	For an ordered structure $\A$ and an FO assignment $v$ for $\A$, $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$ if, and only if, there exists an SO assignment $V$ for $\A$ such that $(\A,V,v)\models \varphi_i'(\bar{X},\bar{x}, \bar{y})$.
\end{clm}
\proof
	To prove the {\em only if} direction, let $\A$ be an ordered structure with domain $A$ and let $v$ be a first-order assignment for $\A$, such that $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$.
	Define $\bar{B} = (B_1,\ldots,B_{\length{\bar{X}}})$ as $B_j = \{v(\bar{w})\mid \text{ $X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$, and let $V$ be a second-order assignment for which $V(\bar{X}) = \bar{B}$.
	Towards a contradiction, suppose that $(\A,V,v)\not\models \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$.
	By the choice of $v$, and construction of $V$ it is clear that $(\A,V,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})\wedge\varphi^{+}_i(\bar{X},\bar{y})$, so we necessarily have that $(\A,V,v)\not\models\varphi^{-}_i(\bar{X},\bar{y})$.
	Let $X_t$ be such that $\neg X_t(\bar{w})$ is mentioned in $\varphi^{-}_i(\bar{X},\bar{y})$ and $(\A,V, v)\not\models\neg X_t(\bar{w})$, namely, $v(\bar{w})\in B_t$. 
	However, by the construction of $B_t$, there exists a subtuple $\bar{z}$ of $\bar{y}$ such that $X_t(\bar{z})$ appears in $\varphi^{+}_i(\bar{X},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$. Since $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$, then the weak ordering in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$. This violates condition (c) above since $\neg X_t(\bar{w})$ appears in $\varphi^{-}_i$ and $X_t(\bar{z})$ appears in $\varphi^{+}_i$, which leads to a contradiction.

@@

The previous claim its subsequent proof motivate the following definition. For a first-order assignment $v$ over $\A$, define $\bar{B}^v = (B^v_1,\ldots,B^v_{\length{\bar{X}}})$ where each $B^v_j = \{v(\bar{w}) \mid \text{$X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$.
One can easily check that for every assignment $(V, v)$ such that $(\A,V,v)\models \varphi_i'(\bar{X}, \bar{x}, \bar{y})$, it holds that $(\A,\bar{B}^v,v)\models \varphi_i'(\bar{X}, \bar{x}, \bar{y})$ and $\bar{B}^v \subseteq V(\bar{X})$.
Namely, $\bar{B}^v$ is a valid candidate for $\bar{X}$ and, furthermore, it is contained in all satisfying assignments of $\bar{X}$ when $v$ is fixed.
This offers an insight into the main idea of Part (1): by choosing one particular $v$ the plan is to remove $\bar{B}^v$ as an assignment over $\bar{X}$ in $\varphi_i$. 
For this, we choose the minimal $v$ that satisfies $\varphi^{\fo}_i(\bar{x},\bar{y})$ which can be defined with the following formula:
\[
\text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \; = \;  \,\varphi_i^{\fo}(\bar{x},\bar{y})\wedge \fa{\bar{x}'} \fa{\bar{y}'}\big(\varphi_i^{\fo}(\bar{x}',\bar{y}')\to ( \bar{x}\leq\bar{x}' \wedge  \bar{y}\leq\bar{y}')\big).
\] 
If $\varphi^{\fo}_i$ is satisfiable, let $v$ be the only assignment such that  $(\A,v)\models \text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y})$. 
Furthermore, let $V^*$ be the second-order assignment and $v^*$ the first-order assignment that satisfy $V^*(\bar{X}) = \bar{B}^{v}$ and $v^*(\bar{x}) = v(\bar{x})$.
By the previous discussion, $(\A,V^*,v^*)\models\varphi_i(\bar{X},\bar{x})$.
